### 2021-07-28 paper/localpsf.tex
edits to impulse response regular grid discussion

We build upon the class of product convolution approximations in which the functions, $\varphi_i$, are impulse responses of $\Aop$ to point sources at a collection of points $x_i$ \cite{PRODCONVLIST}. A popular choice for methods in this class is to choose the points $x_i$ to be nodes in a regular grid, and interpolate the functions $\varphi_i$ with piecewise linear interpolation \cite{NAGY}, or splines \cite{PRODCONVSPLINES}. With a regular grid, a large number of points $x_i$ is typically required to achieve an accurate product-convolution approximation. This is computationally prohibitive in our applications, because computing the functions $\varphi_i$ requires one matrix vector product with $\Aop$ for each point $x_i$, and these matrix vector products are computationally costly. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. But even with adaptive refinement, many matrix vector products with $\Aop$ may be required. In this paper, rather than using adaptive refinement, we instead reduce the computational cost by picking points $x_i$ such that many $\varphi_i$ are computed with each matrix vector product.

@@

To briefly summarize, we extract $\varphi_{x_i}$ for many points $x_i$ by applying $\Aop$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). 
%To reduce memory usage, all kernels $\varphi_i$ associated with a given batch, $b$, are stored in a single function, $\eta^b$, rather than as distinct functions. 
A greedy algorithm is used to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\phi_x$ (Section \ref{eq:mean_and_covariance_estimation}). For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
%\begin{itemize}
%\item We extract $\varphi_{x_i}$ for many points $x_i$ by applying $\mathcal{A}$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). To reduce memory usage, all kernels $\varphi_i$ associated with a given batch, $b$, are stored in a single function, $\eta^b$, rather than as distinct functions.
%\item We use a greedy algorithm to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). 
%\item We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\psi_x$ (Section \ref{eq:mean_and_covariance_estimation}). 
%\item For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
%\end{itemize}
The complete algorithm for constructing $\AopPc$ is shown in Algorithm \ref{alg:construct_Atilde}. The method for evaluating kernel entries of $\Phi_\text{pc}$ is shown in Section \ref{sec:eval_matrix_entries}. The method for converting $\AopPc$ to H-matrix format is shown in Section \ref{sec:H_matrix_conversion}. The method for modifying the resulting H-matrix to make it symmetric positive semi-definite is shown in Section \ref{sec:make_spd}.

@@

Because of the non-negative kernel property, for each $x \in \Omega$, the impulse response $\phi_x$ is a non-negative function $y \mapsto \phi_x(y)$. Hence, $\phi_x$ is a scaled probability distribution. Let $\alpha$ denote the scaling factor, let $\widehat{\phi}_x := \phi_x / \alpha(x)$ denote the normalized version of $\phi_x$, and let $\mu(x)$ and $\Sigma(x)$ denote the mean and covariance of $\widehat{\phi}_x$, respectively. We make the approximation that the support of $\phi_x$ is contained within the ellipsoid
\begin{equation}
\label{eq:support_ellipsoid}
E_x := \{x' \in \Omega: (x' - \mu(x))^T \Sigma(x)^{-1} (x' - \mu(x)) \le \tau^2\},
\end{equation}
where $\tau$ is a fixed constant. 
%The ellipsoid $E_{x}$ is the set of points within $\tau$ standard deviations of the mean of the Gaussian approximation of the normalized version of $\langle \mathcal{A}, \delta_{x}\rangle^{*}$ which has the same mean and covariance. 
The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations from the mean of the Gaussian distribution with mean $\mu(x)$ and covariance $\Sigma(x)$, i.e., the Gaussian distribution which has the same mean and covariance as $\widehat{\phi}_x$ (see Figure FIG). The quantity $\tau$ is a parameter that must be chosen by the user of the algorithm. The larger $\tau$ is, the larger the ellipsoid $E_x$ is, and the more conservative the estimate is for the support of $\phi_x$. However, the cost of our algorithm will depend on how many non-overlapping ellipsoids $E_x$ we can ``pack'' in the domain $\Omega$ (more ellipsoids $\implies$ lower cost), and choosing a larger value of $\tau$ means that fewer ellipsoids will fit in $\Omega$. We find that $\tau=3$ yields a reasonable balance between these competing interests, and use $\tau=3$ in our numerical results. The fraction of the ``mass'' of $\phi_x$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though this bound is conservative and typically far less mass resides in this region. 
% a good balance between these competing goals. 
%The bigger $\tau$ is, the smaller the overlap between impulse responses within a batch will be, while the smaller $\tau$ is, the more impulse responses we can fit within each batch.

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\alpha(x)$, the length-$d$ vector $\mu(x)$, and the $d \times d$ matrix $\Sigma(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$ is to apply $\Aop$ to a delta distribution centered at point $x$ to get $\phi_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\phi_x$ to determine $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$. But this direct approach is computationally infeasible because and it requires applying $\Aop$ to one vector per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\alpha$, $\mu$, and $\Sigma$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

@@

\subsection{Computation and storage of impulse responses}
\label{sec:get_impulse_response}

We compute impulse responses of $\mathcal{A}$ in batches by applying $\mathcal{A}$ to a small number Dirac combs. 
%The batches of impulse responses, $\eta^b$, will ultimately be interpolated with weighting functions described in Section \ref{sec:weighting_functions} to form the desired approximation $\widetilde{\mathcal{A}}\approx \mathcal{A}$.  
The Dirac comb, $\xi_b$, associated with a batch of sample points, $S_b$, is the sum of Dirac distributions (point sources) centered at the points $x_i \in S_b$. That is,
\begin{equation*}
	\xi^b := \sum_{x_i \in S_b} \delta_{x_i}.
\end{equation*}
For each batch $S_b$, we compute the action of $\mathcal{A}$ on the associated Dirac comb:
\begin{equation}
\label{eq:dirac_comb_H_action}
	\eta^b := \langle\mathcal{A}, \xi^b\rangle^*.
\end{equation}
By linearity, the resulting function $\eta^b$ may be written as
\begin{equation}
\label{eq:phi_b}
	\eta^b = \left\langle\mathcal{A},\sum_{x_i \in S_b} \delta_{x_i}\right\rangle^* = \sum_{x_i \in S_b} \langle\mathcal{A}, \delta_{x_i}\rangle^* = \sum_{x_i \in S_b} \phi_x.
\end{equation}
Since the points $x_i$ are chosen so that the support of $\phi_i$ and the support of $\phi_j$ do not overlap (or do not overlap much) when $i \neq j$, from \eqref{eq:convolution_kernel} and \eqref{eq:phi_b} we have
\begin{equation}
\label{eq:varphi_eval}
	\varphi_{x_i}(z) = \phi_i(z+x_i) = \begin{cases}
		\eta^b(z+x_i), & z+x_i \in E_{x_i} \\
		0, & \text{otherwise}.
		\end{cases}
\end{equation}
for all $x_i \in S_b$. By performing one matrix-vector product, \eqref{eq:dirac_comb_H_action}, we recover $\varphi_{x_i}$ for every point $x_i \in S_b$.

Later we will need to perform function evaluations $z \mapsto \varphi_{x_i}(z)$ for a large number of points $z$. Using \eqref{eq:varphi_eval} to perform this task requires evaluating the function $\eta_b$ at a large number of points $z+x_i$. In computations the functions $\eta^b$ will be finite element functions formed on an irregular mesh, and the points $z+x_i$ will typically not be gridpoints of the mesh. Evaluating a finite element function at an arbitrary point requires finding which cell in the mesh that the point is in, and then evaluating several polynomial functions to evaluate the finite element function at the point in that cell. While this procedure is asymptotically scalable, it tends to be slow in practice \cite{FEMEVALSLOW}. To speed up evaluation of the function $\varphi_{x_i}$, we interpolate $\varphi_{x_i}$ onto a regular rectilinear grid on a coordinate axis aligned box that just barely contains the ellipsoid $E_{x_i}$. The function $\varphi_{x_i}$ is transferred from the irregular mesh to the regular rectilinear grid by evaluating $\varphi_{x_i}$ at all gridpoints in the regular rectilinear grid using \eqref{eq:varphi_eval} and the slow finite element evaluation process described previously. Once $\varphi_{x_i}$ has been transferred to the regular rectilinear grid, further evaluations of $\varphi_{x_i}$ are performed using fast regular rectilinear grid interpolation. To limit the error associated with transferring the function $\varphi_{x_i}$ from the irregular mesh to the regular rectilinear grid, the density of the regular rectilinear grid should be denser than the local density of the finite element mesh. In our numerical results, the distance between neighboring points in the regular rectilinear grid is chosen to be half the minimum distance between any pair of finite element Lagrange nodes that are contained in the ellipsoid $E_{x_i}$. This grid transfer procedure is illustrated in Figure FIG.

@@

In this section we show how to construct an H-matrix representation of our product-convolution approximation, after discretizing the problem with finite elements. Let $f_1, f_2, \dots, f_N$ be a set of finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, and let $V_h := \Span\left(f_1, f_2, \dots, f_N\right)$ be the corresponding finite element space. Further, let $p_i \in \mathbb{R}^d$, $i=1,\dots, N$ be the Lagrange nodes associated with the functions $f_i$. That is, the points such that $f_i(p_j)$ equals one if $i=j$, and zero otherwise.

In Section \ref{sec:intro} we noted that the original integral kernel, $\Aker$, is inaccessible, while in Equation \eqref{eq:kernel_entries}  we saw that the integral kernel for our product-convolution approximation, $\AkerPc \approx \Aker$, is accessible, and may be cheaply evaluated at arbitrary pairs of points $(y,x) \in \Omega \times \Omega$ via the following formula:
\begin{equation}
\label{eq:eval_y_x}
\AkerPc(y,x) = \sum_{i=1}^r w_i(x) \varphi_{x_i}(y-x).
\end{equation}
We now define a further approximation, $\AkerPcMesh \approx \AkerPc$, as follows:
\begin{equation}
\label{eq:defn_of_Akerpcmesh}
\AkerPcMesh(y,x) := \sum_{i=1}^N \sum_{j=1}^N \AkerPc(p_i,p_j) f_i(y) f_j(x)
\end{equation}
The kernel $\AkerPcMesh$ is the interpolation of $\AkerPc$ onto $V_h \otimes V_h$. Error in the overall kernel approximation, $\AkerPcMesh \approx \Aker$, arises both because of error in the product-convolution approximation, and because of finite element discretization error incurred from interpolation onto $V_h \otimes V_h$.

@@

We form an H-matrix representation of $\AmatPc$ by converting $\AkerPcMat$ and $\mathbf{M}$ to H-matrix format, then using fast H-matrix methods to multiply these matrices according \eqref{eq:Amatpc_defn}. Once $\AmatPc$ is in H-matrix format, useful matrix operations such as matrix inversion may be performed quickly using fast H-matrix methods. We use H1 matrices in our numerical results, but any of the other H-matrix formats (such as H2, HODLR, HSS, HBS, and others \cite{HMATRIX}) could be used instead.

We convert $\AkerPcMat$ into H1 matrix format using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package \cite{HLIBPRO}. Briefly, this method uses geometric information about the spatial locations of the degrees of freedom associated with the rows and columns of the matrix (here, these locations are the finite element Lagrange nodes $p_i$, $i=1,\dots,N$) to construct a recursive block partitioning of $\AkerPcMat$, such that all blocks at the finest level of the partitioning are either expected to be low rank, or are small. Then, the method forms low rank approximations of the blocks that are expected to be low rank by sampling specially chosen rows and columns of those blocks. The small high rank blocks are formed directly by evaluating their entries. We describe the H-matrix construction in detail in Appendix \ref{app:h_matrix}. Although $\AkerPcMat$ is dense, constructing the H-matrix approximation via this process only requires evaluation of $O(N \log N)$ entries of $\AkerPc$, and these entries are computed via the formula in Equation \eqref{eq:eval_y_x}. We convert $\mathbf{M}$ to H-matrix format using standard H-matrix methods for sparse matrices implemented within HLIBpro, using the same recursive block partitioning structure as was used for $\AkerPcMat$.

@@

\section{Hierarchical matrix construction}
\label{app:h_matrix}

The H-matrix construction process that we use proceeds as follows. First, we construct hierarchical partitionings of the degrees of freedom for the columns and rows of the matrix (cluster trees, Section \ref{sec:cluster_trees}). Second, we construct a hierarchical partitioning of the blocks of the matrix, in such a way that many of the blocks in the partitioning are expected to be low rank, and the remaining high rank blocks are small (block cluster tree, Section \ref{sec:block_cluster_tree}). Finally, we form low rank approximations of the blocks of the matrix that are expected to be low rank (adaptive cross approximation, Section \ref{sec:adaptive_cross}), and fill in the remaining high rank  blocks with their numerical values. The first two steps require geometric information about the spatial locations of the degrees of freedom associated with the rows and columns of the matrix, but these steps do not depend on the particular values of matrix entries. The third step requires us to evaluate $O(N \log N)$ specially-chosen entries of $\widetilde{A}$, and we evaluate these entries using the method described in Section \ref{sec:eval_matrix_entries}.

>>>>

We build upon the class of product convolution approximations in which the functions, $\varphi_i$, are impulse responses of $\Aop$ to point sources at a collection of points $x_i$ \cite{PRODCONVLIST}. A popular choice for methods in this class is to choose the points $x_i$ to be nodes in a regular grid, and interpolate the functions $\varphi_i$ with piecewise linear interpolation \cite{NAGY}, or splines \cite{PRODCONVSPLINES}. With a regular grid, a large number of points $x_i$ is typically required to achieve an accurate product-convolution approximation.
%This is computationally prohibitive in our applications, because computing the functions $\varphi_i$ requires one matrix vector product with $\Aop$ for each point $x_i$, and these matrix vector products are computationally costly. 
In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. But even with adaptive refinement, many matrix vector products with $\Aop$ may be required. In this paper, rather than using adaptive refinement, we instead reduce the computational cost by picking points $x_i$ such that many $\varphi_i$ are computed with each matrix vector product.

@@

To briefly summarize, we extract $\varphi_{x_i}$ for many points $x_i$ by applying $\Aop$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). 
%To reduce memory usage, all kernels $\varphi_i$ associated with a given batch, $b$, are stored in a single function, $\eta^b$, rather than as distinct functions. 
A greedy algorithm is used to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\phi_x$ (Section \ref{eq:mean_and_covariance_estimation}). The support of $\phi_x$ is estimated to be contained within an ellipsoid, so the process of choosing batches of points is an ellipse packing problem. For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
%\begin{itemize}
%\item We extract $\varphi_{x_i}$ for many points $x_i$ by applying $\mathcal{A}$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). To reduce memory usage, all kernels $\varphi_i$ associated with a given batch, $b$, are stored in a single function, $\eta^b$, rather than as distinct functions.
%\item We use a greedy algorithm to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). 
%\item We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\psi_x$ (Section \ref{eq:mean_and_covariance_estimation}). 
%\item For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
%\end{itemize}
The complete algorithm for constructing $\AopPc$ is shown in Algorithm \ref{alg:construct_Atilde}. The method for evaluating kernel entries of $\Phi_\text{pc}$ is shown in Section \ref{sec:eval_matrix_entries}. The method for converting $\AopPc$ to H-matrix format is shown in Section \ref{sec:H_matrix_conversion}. The method for modifying the resulting H-matrix to make it symmetric positive semi-definite is shown in Section \ref{sec:make_spd}.

@@

Because of the non-negative kernel property, for each $x \in \Omega$, the impulse response $\phi_x$ is a non-negative function $y \mapsto \phi_x(y)$. Hence, $\phi_x$ is a scaled probability distribution. Let $\alpha$ denote the scaling factor, let $\widehat{\phi}_x := \phi_x / \alpha(x)$ denote the normalized version of $\phi_x$, and let $\mu(x)$ and $\Sigma(x)$ denote the mean and covariance of $\widehat{\phi}_x$, respectively. We make the approximation that the support of $\phi_x$ is contained within the ellipsoid
\begin{equation}
\label{eq:support_ellipsoid}
E_x := \{x' \in \Omega: (x' - \mu(x))^T \Sigma(x)^{-1} (x' - \mu(x)) \le \tau^2\},
\end{equation}
where $\tau$ is a fixed constant. 
%The ellipsoid $E_{x}$ is the set of points within $\tau$ standard deviations of the mean of the Gaussian approximation of the normalized version of $\langle \mathcal{A}, \delta_{x}\rangle^{*}$ which has the same mean and covariance. 
The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations from the mean of the Gaussian distribution with mean $\mu(x)$ and covariance $\Sigma(x)$, i.e., the Gaussian distribution which has the same mean and covariance as $\widehat{\phi}_x$ (see Figure FIG). The quantity $\tau$ is a parameter that must be chosen by the user of the algorithm. The larger $\tau$ is, the larger the ellipsoid $E_x$ is, the more conservative the estimate is for the support of $\phi_x$. However, the cost of our algorithm will depend on how many non-overlapping ellipsoids $E_x$ we can ``pack'' in the domain $\Omega$ (more ellipsoids $\implies$ lower cost), and choosing a larger value of $\tau$ means that fewer ellipsoids will fit in $\Omega$. We find that $\tau=3$ yields a reasonable balance between these competing interests, and use $\tau=3$ in our numerical results. The fraction of the ``mass'' of $\phi_x$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though this bound is conservative and typically far less mass resides in this region. 
% a good balance between these competing goals. 
%The bigger $\tau$ is, the smaller the overlap between impulse responses within a batch will be, while the smaller $\tau$ is, the more impulse responses we can fit within each batch.

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\alpha(x)$, the length-$d$ vector $\mu(x)$, and the $d \times d$ matrix $\Sigma(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\phi_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\phi_x$ to determine $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$. But this direct approach is computationally infeasible because and it requires applying $\Aop$ to one vector per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\alpha$, $\mu$, and $\Sigma$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

@@

\subsection{Computation, evaluation, and storage of impulse responses}
\label{sec:get_impulse_response}

We compute impulse responses of $\mathcal{A}$ in batches by applying $\mathcal{A}$ to a small number Dirac combs. Each Dirac comb yields one batch of impulse responses.
%The batches of impulse responses, $\eta^b$, will ultimately be interpolated with weighting functions described in Section \ref{sec:weighting_functions} to form the desired approximation $\widetilde{\mathcal{A}}\approx \mathcal{A}$.  
The Dirac comb, $\xi_b$, associated with a batch of sample points, $S_b$, is the sum of Dirac distributions (point sources) centered at the points $x_i \in S_b$. That is,
\begin{equation*}
	\xi^b := \sum_{x_i \in S_b} \delta_{x_i}.
\end{equation*}
For each batch $S_b$, we compute the action of $\mathcal{A}$ on the associated Dirac comb:
\begin{equation}
\label{eq:dirac_comb_H_action}
	\eta^b := \langle\mathcal{A}, \xi^b\rangle^*.
\end{equation}
By linearity, the resulting function $\eta^b$ may be written as
\begin{equation}
\label{eq:phi_b}
	\eta^b = \left\langle\mathcal{A},\sum_{x_i \in S_b} \delta_{x_i}\right\rangle^* = \sum_{x_i \in S_b} \langle\mathcal{A}, \delta_{x_i}\rangle^* = \sum_{x_i \in S_b} \phi_x.
\end{equation}
Since the points $x_i$ are chosen so that the ellipsoid $E_{x_i}$ that (approximately) supports $\phi_i$, and the ellipsoid $E_{x_j}$ that (approximately) supports $\phi_j$ do not overlap when $i \neq j$, from \eqref{eq:convolution_kernel} and \eqref{eq:phi_b} we have
\begin{equation}
\label{eq:varphi_eval}
	\varphi_{x_i}(z) = \phi_i(z+x_i) = \begin{cases}
		\eta^b(z+x_i), & z+x_i \in E_{x_i} \\
		0, & \text{otherwise}.
		\end{cases}
\end{equation}
for all $x_i \in S_b$. By performing one matrix-vector product, \eqref{eq:dirac_comb_H_action}, we recover the functions $\varphi_{x_i}$ for every point $x_i \in S_b$.

To convert $\AopPc$ to H-matrix format, we will need to evaluate the function $z \mapsto \varphi_{x_i}(z)$ for a large number of points $z$. Using \eqref{eq:varphi_eval} to evaluate $\varphi_{x_i}(z)$ requires evaluating the function $\eta_b(z+x_i)$. However, in computations the function $\eta^b$ will be a finite element function formed on an irregular mesh, and the point $z+x_i$ will typically not be a gridpoint of the mesh. Evaluating a finite element function at an arbitrary point requires finding which cell in the mesh the point is in, then evaluating several polynomial functions. This procedure is asymptotically scalable, but slow in practice \cite{FEMEVALSLOW}. To speed up evaluation of $\varphi_{x_i}$, we interpolate $\varphi_{x_i}$ onto a regular rectilinear grid on a coordinate axis aligned box that just barely contains the ellipsoid $E_{x_i}$. We transfer the function $\varphi_{x_i}$ from the irregular mesh to the regular grid by evaluating $\varphi_{x_i}$ at all gridpoints in the regular grid, using \eqref{eq:varphi_eval} and the slow finite element point evaluation process described previously. Once $\varphi_{x_i}$ has been transferred to the regular grid, we use fast regular grid interpolation to perform further evaluations of $\varphi_{x_i}$. We incur interpolation error when transferring $\varphi_{x_i}$ from the irregular mesh to the regular grid. To limit this interpolation error, the density of the regular grid should be denser than the local density of the finite element mesh. In our numerical results, the distance between neighboring points in the regular grid is chosen to be half the minimum distance between any pair of finite element Lagrange nodes that are contained in $E_{x_i}$. This grid transfer procedure is illustrated in Figure FIG.

@@

In this section we discuss how we construct an H-matrix representation of our product-convolution approximation, after discretizing the problem with finite elements. Let $f_1, f_2, \dots, f_N$ be a set of finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, and let $V_h := \Span\left(f_1, f_2, \dots, f_N\right)$ be the corresponding finite element space. Further, let $p_i \in \mathbb{R}^d$, $i=1,\dots, N$ be the Lagrange nodes associated with the functions $f_i$. That is, the points such that $f_i(p_j)$ equals one if $i=j$, and zero otherwise.

In Section \ref{sec:intro} we noted that the original integral kernel, $\Aker$, is inaccessible, while in Equation \eqref{eq:kernel_entries}  we saw that the integral kernel for our product-convolution approximation, $\AkerPc \approx \Aker$, is accessible, and may be cheaply evaluated at arbitrary pairs of points $(y,x) \in \Omega \times \Omega$ via the following formula:
\begin{equation}
\label{eq:eval_y_x}
\AkerPc(y,x) = \sum_{i=1}^r w_i(x) \varphi_{x_i}(y-x).
\end{equation}
We now define a further approximation, $\AkerPcMesh \approx \AkerPc$, as follows:
\begin{equation}
\label{eq:defn_of_Akerpcmesh}
\AkerPcMesh(y,x) := \sum_{i=1}^N \sum_{j=1}^N \AkerPc(p_i,p_j) f_i(y) f_j(x).
\end{equation}
The kernel $\AkerPcMesh$ is the interpolation of $\AkerPc$ onto $V_h \otimes V_h$. Error in the overall kernel approximation, $\AkerPcMesh \approx \Aker$, arises both because of error in the product-convolution approximation, and because of finite element discretization error incurred from interpolation onto $V_h \otimes V_h$.

@@

We form an H-matrix representation of $\AmatPc$ by converting $\AkerPcMat$ and $\mathbf{M}$ to H-matrix format, then using fast H-matrix methods to multiply these matrices according \eqref{eq:Amatpc_defn}. Once $\AmatPc$ is in H-matrix format, useful matrix operations such as matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, and matrix inversion may be performed using fast scalable H-matrix methods. We use H1 matrices in our numerical results, but any of the other H-matrix formats (such as H2, HODLR, HSS, HBS, and others \cite{HMATRIX}) could be used instead. For more details on H-matrices, we recommend \cite{HMATRIXGOOD}. We discuss H-matrices in greater detail in Appendix \ref{app:h_matrix}

We convert $\AkerPcMat$ into H1 matrix format using the standard geometrical clustering/adaptive cross method implemented within the HLIBPro software package \cite{HLIBPRO}. Briefly, this method uses geometric information about the spatial locations of the degrees of freedom associated with the rows and columns of the matrix (here, these locations are the finite element Lagrange nodes $p_i$, $i=1,\dots,N$) to construct a recursive block partitioning of $\AkerPcMat$, such that all blocks at the finest level of the partitioning are either expected to be low rank, or are small. Then, the method forms low rank approximations of the blocks that are expected to be low rank by sampling specially chosen rows and columns of those blocks. The small high rank blocks are formed directly by evaluating their entries. We describe this H-matrix construction process in detail in Appendix \ref{app:h_matrix}. Although $\AkerPcMat$ is a dense $N \times N$ matrix, constructing the H-matrix approximation of $\AkerPcMat$ only requires evaluation of $O(N \log N)$ entries of $\AkerPc$, and these entries are computed via the formula in Equation \eqref{eq:eval_y_x}. A dense representation of $\AkerPcMat$ is never formed. We convert $\mathbf{M}$ to H-matrix format using standard H-matrix methods for sparse matrices implemented within HLIBPro, using the same recursive block partitioning structure as was used for $\AkerPcMat$.

@@

\section{Hierarchical matrix details}
\label{app:h_matrix}

The H-matrix construction process that we use to form the matrix $\AkerPcMat$ proceeds as follows. First, we construct hierarchical partitionings of the degrees of freedom for the columns and rows of the matrix (cluster trees, Section \ref{sec:cluster_trees}). Second, we construct a hierarchical partitioning of the blocks of the matrix, in such a way that many of the blocks in the partitioning are expected to be low rank, and the remaining high rank blocks are small (block cluster tree, Section \ref{sec:block_cluster_tree}). Finally, we form low rank approximations of the blocks of the matrix that are expected to be low rank (adaptive cross approximation, Section \ref{sec:adaptive_cross}), and fill in the remaining high rank  blocks with their numerical values. The first two steps require geometric information about the spatial locations of the degrees of freedom associated with the rows and columns of the matrix, but these steps do not depend on the particular values of matrix entries. The third step requires us to evaluate $O(N \log N)$ specially-chosen entries of $\widetilde{A}$, and we evaluate these entries using the method described in Section \ref{sec:eval_matrix_entries}.
